### 2016-09-18 paper.tex
Correct fig ref format; resolves #77

Geometric parameters for the 1:6 scale RM2 rotor were taken from the RM2 design
report \cite{Barone2011}, with the exception of the shaft diameter, which was
scaled from the SAFL RM2 model \cite{Hill2014}. Values for both the 1:6 and
full-scale designs are presented in Table~\ref{tab:turb-geom} and a drawing of
the turbine is shown in Fig.~\ref{fig:turbine-drawing}. The rotor
components---blades, struts, shaft, and center hub sections---were fabricated
from 6061-T6 aluminum, which was hardcoat anodized per MIL-8625-A, type III,
class 2 specifications. CAD models of the turbine are available from
\cite{Bachant2015-RM2-CAD}.

@@

For investigating the effects of support strut drag losses, a set of cylindrical
covers were designed to slip over the struts, which provided a deliberate and
dramatic increase in strut drag. Endcaps were also fabricated to allow the high
drag strut cover configuration to be operated without blades. A drawing of the
strut covers is shown in Fig.~\ref{fig:covers}.

@@

Measurements were conducted in the UNH tow/wave tank, pictured in
Fig.~\ref{fig:exp-setup-photo}, a 36 m long facility with a 3.66 m wide by 2.44
m deep cross-section, capable of tow speeds up to 3 m/s. The blockage ratio
based on the rotor frontal and tank cross-sectional area was 10\%. The turbine
was mounted in a frame built from NACA 0020 struts, attached to the tow carriage
by four linear bearings, which transfer all streamwise force to a pair of 500
lbf capacity Sentran ZB3 S-beam load cells \cite{SentranZB}. The turbine shaft
RPM was controlled by a servo motor system, which allowed precise control of the
turbine tip speed ratio. The load torque was measured by an Interface T8 200 Nm
capacity inline rotary torque transducer \cite{InterfaceT8}. An additional
Sentran ZB3 load cell mounted at a fixed distance from the servo motor via a
moment arm provided a redundant torque measurement, measurements from which were
nearly identical to the inline transducer, providing additional confidence in
the results. Turbine shaft angle was measured using the Kollmorgen AKD servo
drive's emulated encoder output \cite{KollmorgenAKD}, set to $10^5$ counts per
turbine shaft revolution. Carriage speed, and therefore inflow velocity was
measured using a Renishaw LM15 linear encoder with 10 $\mu$m resolution
\cite{RenishawLM15}. All of these performance-related quantities were sampled at
2 kHz, while the tow tank's motion controller provided redundant measurements of
the carriage speed turbine angular velocity sampled at 1 kHz. Turbine wake
measurements at 1 meter downstream were measured with a Nortek Vectrino+
acoustic Doppler velocimeter \cite{NortekVectrino}, sampling at 200 Hz. A photo
and drawing of the experimental setup are shown in
Fig.~\ref{fig:exp-setup-photo} and Fig.~\ref{fig:exp-setup}, respectively.

@@

Wake measurements were all performed at 1 m downstream, which corresponds to
$x/D = 0.93$. The cross-stream and vertical coordinates are shown in
Fig.~\ref{fig:coordinates}. Altogether 750 tows were performed and included in
the experimental database.

@@

Mean rotor power coefficients for multiple Reynolds numbers are plotted versus
tip speed ratio in Fig.~\ref{fig:cp-curves}. In general, $C_P$ increases with
$Re$, along with a reduction in the optimal tip speed ratio, due to the tendency
of foils to stall at higher angles of attack at higher $Re$, and the higher
angle of attack ranges seen by the blades at lower $\lambda$. These effects
diminish with increasing $Re$, which is expected as the blade boundary layers
transition to turbulence closer to the leading edge \cite{Lissaman1983,
    McMasters1980, Bachant2016-RVAT-Re-dep}, which helps flow remain attached longer
as it moves against the adverse pressure gradient on the suction side of the
foil. For the experiment reported here, a maximum power coefficient $C_P=0.37$
was reached at $Re_D=1.3 \times 10^6$. Note how at lower tip speed ratios and
Reynolds numbers, $C_P$ becomes significantly negative, which was possible with
the speed control of the experimental setup's servo motor applying negative
torque to the rotor.

@@

Mean rotor drag (or thrust) coefficients are plotted versus tip speed ratio in
Fig.~\ref{fig:cd-curves}. These $C_D$ curves show little difference compared to
the power coefficient curves; differences are most noticeable at low $\lambda$
and low $Re$. The relative similarity could be attributed to the effects of
stall, where the lift-to-drag ratio on the blades may drop, decreasing rotor
torque, though the total resultant force due to high blade drag retains a
similar component in the streamwise direction. When operating at maximum power
coefficient, $\lambda_0=3.1$, and $Re_D=1.3 \times 10^6$, a rotor drag
coefficient $C_D=0.84$ was measured.

@@

The effects of Reynolds number on the power and drag coefficients at
$\lambda=3.1$ are shown in Fig.~\ref{fig:perf-re-dep}. The rotor drag
coefficient $C_D$ became more or less Reynolds number independent for $Re_D \ge
0.6 \times 10^6$. The power coefficient of the turbine increased dramatically
below $Re_D = 1 \times 10^6$ or $Re_c \equiv \lambda U_\infty c / \nu \approx 2
\times 10^5$, beyond which there appears to be a small, linear, positive trend.
At the lowest Reynolds number, mean power coefficient even dropped below zero,
which is consistent with the low performance of the 1:15 scale RM2 physical
model study~\cite{Hill2014}. However, the blades of the 1:15 scale RM2 were
mounted at approximately 58\% of the chord from the leading edge, versus 50\%
for the present model, which helps explain the 1:15 scale model's positive power
output at $\lambda=2.2$ and $Re_D \sim 10^5$, since moving the blade mount point
further back is equivalent to a ``toe-out'' preset pitch
condition~\cite{Fiedler2009}.

The tendency of $C_P$ to continue increasing slightly could be an effect of flow
curvature---caused by the finite $c/R$---which imparts a ``virtual camber''
\cite{Migliore1980}, or can be thought of as producing a ``lead'' in angle of
attack since cambered airfoils have non-zero lift at zero angle of attack
\cite{Goude2012}. The residual weak $Re$-dependence of the RM2 compared to the
$Re$-independence of the higher $c/R$ UNH-RVAT, shown in
Fig.~\ref{fig:rvat-perf-re-dep}, could be due to this virtual camber effect,
since camber has been shown to cause earlier $Re$-convergence of the CFT's
geometric torque coefficient when calculated from static foil coefficients given
by a viscous panel method, whereas characteristics like lift-to-drag ratio do
not converge as strongly \cite{Bachant2016-Energies}.

@@

Performance curves for the rotor with both NACA 0021 and cylindrical struts are
shown in Fig.~\ref{fig:perf-covers}. As expected, the high drag cylindrical
struts reduce performance dramatically, producing an entirely negative $C_P$
curve. The detrimental effects of the strut drag were more pronounced at higher
$\lambda$. This is in accordance with the measurements of Rawlings
\cite{Rawlings2008}, though the strut losses in this case were larger due to the
very high drag circular profile. In contrast to the dramatic effect strut drag
had on $C_P$, overall rotor drag measurements were relatively similar for both
the cylindrical and NACA 0021 strut cases.

@@

Measurements for the power coefficient contributions of the strut drag losses
are presented in Fig.~\ref{fig:no-blades} for NACA 0021 and cylindrical
struts---both in towed (at $U_\infty=1$ m/s) and stationary conditions. These
are computed in the same fashion as the curves in Fig.~\ref{fig:cp-curves}, but
with the rotor blades removed. For the stationary case, tip speed ratio and
power coefficient were computed using the same reference velocity $U_\infty=1$
m/s. We see that strut drag losses increase with tip speed ratio to the power
2--3, which makes streamlined struts much more important for low solidity
turbines, given the inverse correlation between solidity and $\lambda_0$
\cite{Templin1974}.

@@

Strut drag losses did not change much for the streamlined NACA 0021 struts in
the towed versus stationary configuration, which helps explain why overall rotor
drag coefficients shown in Fig.~\ref{fig:perf-covers} remained of comparable
magnitude even at low Reynolds number, where $C_P$ was dramatically reduced, or
even negative. For the cylindrical struts, losses increased significantly when
towed and operating in the mid range of tip speed ratios.

@@

The mean velocity field at the chosen optimal tip speed ratio $\lambda_0=3.1$, 1
m downstream ($x/D=0.93$) from the rotor axis is plotted in
Fig.~\ref{fig:meancontquiv}. The mean streamwise velocity deficit is markedly
more symmetric than that of the higher solidity UNH-RVAT, shown in
Fig.~\ref{fig:rvat-meancontquiv}, with the RM2 inducing lower acceleration
around the turbine due to the lower rotor drag coefficient and a slightly lower
blockage ratio \cite{Bachant2015-JoT}. Tip vortex shedding is relatively weaker,
which is likely an effect of the RM2's smaller blade chord length and tapered
blades.

@@

    \caption{{\bf RM2 Near-Wake Mean Velocity.} RM2 near-wake mean velocity
        field (looking upstream) at 1 m downstream ($x/D=0.93$), $U_\infty=1.0$ m/s,
        and $\lambda=3.1$. Refer to Fig.~\ref{fig:coordinates} for turbine axis
        orientation. Solid dark gray lines indicate turbine frontal area.}

@@

Fig.~\ref{fig:kcont} shows the turbulence kinetic energy in the turbine wake. We
mainly see unsteadiness in the flow generated by the blade tip vortex shedding
(the horizontal band around $z/H=0.5$). Compared with the UNH-RVAT, shown in
Fig.~\ref{fig:rvat-kcont}, turbulence generation is lower overall, without the
intense vertical band around $y/R=-1$, which indicates that the RM2 blades are
operating further from stall---consistent with its higher optimal tip speed
ratio.

@@

Derivatives were computed with a second order central difference scheme for
interior points, and a second order inward-facing scheme for the edges,
following the methodology in \cite{Bachant2015-JoT}. Weighted averages for these
calculations are shown in Fig.~\ref{fig:Ktransport}. Due to the weaker blade
vortex shedding, transport due to vertical advection at this point in the wake
was approximately 3 times lower than the higher solidity UNH-RVAT. Note that
direct comparison is somewhat invalid, since the total measurement plane area
was about 5\% lower for this experiment. However, the differences observed are
larger than the differences in measurement plane area. We also see relatively
lower levels of cross-stream turbulent transport due to the lack of blade stall
vortex shedding. These results may have interesting implications regarding the
application of turbines with lower power coefficient to possibly improve overall
array performance through enhanced transport of kinetic energy from the free
stream, though evaluating these trade-offs will require a detailed analysis of
the downstream evolution and turbine--wake interaction.

>>>>

Geometric parameters for the 1:6 scale RM2 rotor were taken from the RM2 design
report \cite{Barone2011}, with the exception of the shaft diameter, which was
scaled from the SAFL RM2 model \cite{Hill2014}. Values for both the 1:6 and
full-scale designs are presented in Table~\ref{tab:turb-geom} and a drawing of
the turbine is shown in Fig~\ref{fig:turbine-drawing}. The rotor
components---blades, struts, shaft, and center hub sections---were fabricated
from 6061-T6 aluminum, which was hardcoat anodized per MIL-8625-A, type III,
class 2 specifications. CAD models of the turbine are available from
\cite{Bachant2015-RM2-CAD}.

@@

For investigating the effects of support strut drag losses, a set of cylindrical
covers were designed to slip over the struts, which provided a deliberate and
dramatic increase in strut drag. Endcaps were also fabricated to allow the high
drag strut cover configuration to be operated without blades. A drawing of the
strut covers is shown in Fig~\ref{fig:covers}.

@@

Measurements were conducted in the UNH tow/wave tank, pictured in
Fig~\ref{fig:exp-setup-photo}, a 36 m long facility with a 3.66 m wide by 2.44
m deep cross-section, capable of tow speeds up to 3 m/s. The blockage ratio
based on the rotor frontal and tank cross-sectional area was 10\%. The turbine
was mounted in a frame built from NACA 0020 struts, attached to the tow carriage
by four linear bearings, which transfer all streamwise force to a pair of 500
lbf capacity Sentran ZB3 S-beam load cells \cite{SentranZB}. The turbine shaft
RPM was controlled by a servo motor system, which allowed precise control of the
turbine tip speed ratio. The load torque was measured by an Interface T8 200 Nm
capacity inline rotary torque transducer \cite{InterfaceT8}. An additional
Sentran ZB3 load cell mounted at a fixed distance from the servo motor via a
moment arm provided a redundant torque measurement, measurements from which were
nearly identical to the inline transducer, providing additional confidence in
the results. Turbine shaft angle was measured using the Kollmorgen AKD servo
drive's emulated encoder output \cite{KollmorgenAKD}, set to $10^5$ counts per
turbine shaft revolution. Carriage speed, and therefore inflow velocity was
measured using a Renishaw LM15 linear encoder with 10 $\mu$m resolution
\cite{RenishawLM15}. All of these performance-related quantities were sampled at
2 kHz, while the tow tank's motion controller provided redundant measurements of
the carriage speed turbine angular velocity sampled at 1 kHz. Turbine wake
measurements at 1 meter downstream were measured with a Nortek Vectrino+
acoustic Doppler velocimeter \cite{NortekVectrino}, sampling at 200 Hz. A photo
and drawing of the experimental setup are shown in
Figs~\ref{fig:exp-setup-photo}~and~\ref{fig:exp-setup}, respectively.

@@

Wake measurements were all performed at 1 m downstream, which corresponds to
$x/D = 0.93$. The cross-stream and vertical coordinates are shown in
Fig~\ref{fig:coordinates}. Altogether 750 tows were performed and included in
the experimental database.

@@

Mean rotor power coefficients for multiple Reynolds numbers are plotted versus
tip speed ratio in Fig~\ref{fig:cp-curves}. In general, $C_P$ increases with
$Re$, along with a reduction in the optimal tip speed ratio, due to the tendency
of foils to stall at higher angles of attack at higher $Re$, and the higher
angle of attack ranges seen by the blades at lower $\lambda$. These effects
diminish with increasing $Re$, which is expected as the blade boundary layers
transition to turbulence closer to the leading edge \cite{Lissaman1983,
    McMasters1980, Bachant2016-RVAT-Re-dep}, which helps flow remain attached longer
as it moves against the adverse pressure gradient on the suction side of the
foil. For the experiment reported here, a maximum power coefficient $C_P=0.37$
was reached at $Re_D=1.3 \times 10^6$. Note how at lower tip speed ratios and
Reynolds numbers, $C_P$ becomes significantly negative, which was possible with
the speed control of the experimental setup's servo motor applying negative
torque to the rotor.

@@

Mean rotor drag (or thrust) coefficients are plotted versus tip speed ratio in
Fig~\ref{fig:cd-curves}. These $C_D$ curves show little difference compared to
the power coefficient curves; differences are most noticeable at low $\lambda$
and low $Re$. The relative similarity could be attributed to the effects of
stall, where the lift-to-drag ratio on the blades may drop, decreasing rotor
torque, though the total resultant force due to high blade drag retains a
similar component in the streamwise direction. When operating at maximum power
coefficient, $\lambda_0=3.1$, and $Re_D=1.3 \times 10^6$, a rotor drag
coefficient $C_D=0.84$ was measured.

@@

The effects of Reynolds number on the power and drag coefficients at
$\lambda=3.1$ are shown in Fig~\ref{fig:perf-re-dep}. The rotor drag
coefficient $C_D$ became more or less Reynolds number independent for $Re_D \ge
0.6 \times 10^6$. The power coefficient of the turbine increased dramatically
below $Re_D = 1 \times 10^6$ or $Re_c \equiv \lambda U_\infty c / \nu \approx 2
\times 10^5$, beyond which there appears to be a small, linear, positive trend.
At the lowest Reynolds number, mean power coefficient even dropped below zero,
which is consistent with the low performance of the 1:15 scale RM2 physical
model study~\cite{Hill2014}. However, the blades of the 1:15 scale RM2 were
mounted at approximately 58\% of the chord from the leading edge, versus 50\%
for the present model, which helps explain the 1:15 scale model's positive power
output at $\lambda=2.2$ and $Re_D \sim 10^5$, since moving the blade mount point
further back is equivalent to a ``toe-out'' preset pitch
condition~\cite{Fiedler2009}.

The tendency of $C_P$ to continue increasing slightly could be an effect of flow
curvature---caused by the finite $c/R$---which imparts a ``virtual camber''
\cite{Migliore1980}, or can be thought of as producing a ``lead'' in angle of
attack since cambered airfoils have non-zero lift at zero angle of attack
\cite{Goude2012}. The residual weak $Re$-dependence of the RM2 compared to the
$Re$-independence of the higher $c/R$ UNH-RVAT, shown in
Fig~\ref{fig:rvat-perf-re-dep}, could be due to this virtual camber effect,
since camber has been shown to cause earlier $Re$-convergence of the CFT's
geometric torque coefficient when calculated from static foil coefficients given
by a viscous panel method, whereas characteristics like lift-to-drag ratio do
not converge as strongly \cite{Bachant2016-Energies}.

@@

Performance curves for the rotor with both NACA 0021 and cylindrical struts are
shown in Fig~\ref{fig:perf-covers}. As expected, the high drag cylindrical
struts reduce performance dramatically, producing an entirely negative $C_P$
curve. The detrimental effects of the strut drag were more pronounced at higher
$\lambda$. This is in accordance with the measurements of Rawlings
\cite{Rawlings2008}, though the strut losses in this case were larger due to the
very high drag circular profile. In contrast to the dramatic effect strut drag
had on $C_P$, overall rotor drag measurements were relatively similar for both
the cylindrical and NACA 0021 strut cases.

@@

Measurements for the power coefficient contributions of the strut drag losses
are presented in Fig~\ref{fig:no-blades} for NACA 0021 and cylindrical
struts---both in towed (at $U_\infty=1$ m/s) and stationary conditions. These
are computed in the same fashion as the curves in Fig~\ref{fig:cp-curves}, but
with the rotor blades removed. For the stationary case, tip speed ratio and
power coefficient were computed using the same reference velocity $U_\infty=1$
m/s. We see that strut drag losses increase with tip speed ratio to the power
2--3, which makes streamlined struts much more important for low solidity
turbines, given the inverse correlation between solidity and $\lambda_0$
\cite{Templin1974}.

@@

Strut drag losses did not change much for the streamlined NACA 0021 struts in
the towed versus stationary configuration, which helps explain why overall rotor
drag coefficients shown in Fig~\ref{fig:perf-covers} remained of comparable
magnitude even at low Reynolds number, where $C_P$ was dramatically reduced, or
even negative. For the cylindrical struts, losses increased significantly when
towed and operating in the mid range of tip speed ratios.

@@

The mean velocity field at the chosen optimal tip speed ratio $\lambda_0=3.1$, 1
m downstream ($x/D=0.93$) from the rotor axis is plotted in
Fig~\ref{fig:meancontquiv}. The mean streamwise velocity deficit is markedly
more symmetric than that of the higher solidity UNH-RVAT, shown in
Fig~\ref{fig:rvat-meancontquiv}, with the RM2 inducing lower acceleration
around the turbine due to the lower rotor drag coefficient and a slightly lower
blockage ratio \cite{Bachant2015-JoT}. Tip vortex shedding is relatively weaker,
which is likely an effect of the RM2's smaller blade chord length and tapered
blades.

@@

    \caption{{\bf RM2 Near-Wake Mean Velocity.} RM2 near-wake mean velocity
        field (looking upstream) at 1 m downstream ($x/D=0.93$), $U_\infty=1.0$ m/s,
        and $\lambda=3.1$. Refer to Fig~\ref{fig:coordinates} for turbine axis
        orientation. Solid dark gray lines indicate turbine frontal area.}

@@

Fig~\ref{fig:kcont} shows the turbulence kinetic energy in the turbine wake. We
mainly see unsteadiness in the flow generated by the blade tip vortex shedding
(the horizontal band around $z/H=0.5$). Compared with the UNH-RVAT, shown in
Fig~\ref{fig:rvat-kcont}, turbulence generation is lower overall, without the
intense vertical band around $y/R=-1$, which indicates that the RM2 blades are
operating further from stall---consistent with its higher optimal tip speed
ratio.

@@

Derivatives were computed with a second order central difference scheme for
interior points, and a second order inward-facing scheme for the edges,
following the methodology in \cite{Bachant2015-JoT}. Weighted averages for these
calculations are shown in Fig~\ref{fig:Ktransport}. Due to the weaker blade
vortex shedding, transport due to vertical advection at this point in the wake
was approximately 3 times lower than the higher solidity UNH-RVAT. Note that
direct comparison is somewhat invalid, since the total measurement plane area
was about 5\% lower for this experiment. However, the differences observed are
larger than the differences in measurement plane area. We also see relatively
lower levels of cross-stream turbulent transport due to the lack of blade stall
vortex shedding. These results may have interesting implications regarding the
application of turbines with lower power coefficient to possibly improve overall
array performance through enhanced transport of kinetic energy from the free
stream, though evaluating these trade-offs will require a detailed analysis of
the downstream evolution and turbine--wake interaction.
